---
title: "Computational Stats"
site: bookdown::bookdown_site
---

```{r include = FALSE}
PROJHOME <- "/Users/tiago.correia/courses/MestradoBigData/EstatisticaMultivariada"
datasetsDir=file.path(PROJHOME,"datasets")
```


# Lesson 7

## Análise de Discriminantes

```{r}
c1 <- c(-2,0,-1,0,2,1,1,0,-1)
c2 <- c(5,3,1,6,4,2,-2,0,-4)
groups <- c(1,2,3)
G <- rep(groups,each=3)
X <- data.frame(cbind(G,c1,c2))
n <- dim(X)[1]
p <- dim(X)[2]-1
g <- length(groups)

alpha <- 0.05

S <- var(X[,-1])
Slist <- lapply(groups, function(group){
  var(X[X$G==group,-1])
})

nlist <- lapply(groups, function(group){
  dim(X[X$G==group,-1])[1]
})



T <- (n-1)*S

W <- matrix(rep(0,p*p),2)
for(i in 1:length(nlist)){
  W = W + (nlist[[i]]-1)*Slist[[i]]
}

Spooled <- W/(sum(unlist(nlist))-length(nlist))

B <- T - W

eigenDisc <- eigen(solve(W)%*%B)


```

Now we need to understand if the second eigen value is significantly different from 0, and that depends on the variability of the distribution.

For this, we will use the Wilk's lambda test stats:

$$
 - \left(n-1-\frac{p+g}{2}\right) ln \Lambda^* \approx \chi^2_{p(g-1)}
$$

And now we want to test $H_0 : \lambda_i = 0(i = 1,...,s)$, where $s$ is the number of non-null eigen values.
We need to test if each eigen value is significantly different than 0.


Therefore, for the m-esime test we have:

$$
 \Lambda^{*}_{m} = \Pi^{s}_{i=m} \frac{1}{1+l_i}
$$

```{r}

nEigen <- length(eigenDisc$values)

Lw <- sapply(1:nEigen,function(idx){
  prod(1/(1+eigenDisc$values[idx:nEigen]))
})

#Critic Values
CVs <- sapply(1:nEigen,function(idx){
  qchisq(0.95,(p-idx+1)*(g-idx))
})


#\left(n-1-\frac{p+g}{2}\right) ln \Lambda^* \approx \chi^2_{p(g-1)}

LwChisqr <- sapply(Lw,function(currLw){
  -(n-1-((p+g)/2))*log(currLw)
})

Lw > CVs
```

Do we reject any test (for critic regions)?

1. First eigen (`r Lw[1]`) VS critic region (`r CVs[1]`): `r (Lw > CVs)[1]`
2. Second eigen (`r Lw[2]`) VS critic region (`r CVs[2]`): `r (Lw > CVs)[2]`

```{r}
pValues <- sapply(1:nEigen,function(idx){
  1-pchisq(Lw[idx],(p-idx+1)*(g-idx))
})

pValues < alpha

```

Do we reject any test (for p values)?

1. First eigen (`r pValues[1]`) VS critic region (`r alpha`): `r (pValues < alpha)[1]`
2. Second eigen (`r pValues[2]`) VS critic region (`r alpha`): `r (pValues < alpha)[2]`

```{r}

m <- aggregate(X[,2:3],list(X[,1]),mean)

classifications <- apply(X,1,function(obs){
  print(obs)
  obs <- unlist(c(unname(obs[c(2,3)])))
  centered <- obs - as.matrix(m[,-1])
  dists <- diag(centered%*%solve(Spooled)%*%t(centered))
  return(which.min(dists))
})

```

## PCA

```{r}

dados8 <- as.data.frame(readxl::read_xlsx(file.path(datasetsDir, "data8.xlsx")))
dados8.scaled <- scale(dados8, center=TRUE,scale=TRUE)
S <- var(dados8.scaled)

round(S,3)

dados8.eigen <- eigen(S)

eigenAnalysis = data.frame(
  CP=1:length(dados8.eigen$values),
  EV= round(dados8.eigen$values,3),
  Prop_var = round(dados8.eigen$values / length(dados8.eigen$values),3),
  Prop_cum = round(cumsum(dados8.eigen$values / length(dados8.eigen$values)),3)
)

knitr::kable(eigenAnalysis)

plotly::subplot(
  plotly::plot_ly(
    eigenAnalysis, x=~CP, y=~EV, type="scatter", mode="lines+markers", name="Scree Plot"
  ),
  plotly::plot_ly(
    eigenAnalysis, x=~CP, y=~Prop_cum, type="scatter", mode="lines+markers", name="Cum Importance"
  )
)

```

# Lesson 8 

```{r}
dados8 <- as.data.frame(readxl::read_xlsx(file.path(datasetsDir, "data8.xlsx")))
footpca<-prcomp(x = dados8,scale. = TRUE)
summary(footpca)

attributes(footpca)

round(footpca$rotation,3)

screeplot(footpca,type = "l",main="Scree plot")


cbinded_data_pcs <- crosstalk::SharedData$new(cbind.data.frame(as.data.frame(dados8.scaled), as.data.frame(footpca$x)))

crosstalk::bscols(
  d3scatter::d3scatter(cbinded_data_pcs, ~WDIM, ~CIRCUM, color = "blue",width="100%", height=300),
  d3scatter::d3scatter(cbinded_data_pcs, ~PC1, ~PC2, color = "green",width="100%", height=300)
)

```

```{r}
dados8.scaled.df <- as.data.frame(dados8.scaled)

S <- cov(dados8.scaled.df)

S>=0.5 # we have (p^2 - p)/2 distinct correlations, and just 3 of them are >= 0.5

pcor <- corpcor::cor2pcor(S) # this is the parcial correlation between variables excluding the influence of all the others
# the values must be interpreted as the percentage of variability left explained by removing the influence of all the other variables
```

### Aplying the Maunchly Test / Sphericity Test


- We assume that the vector is normal multivariate
- the test stats is 
$$
  U^* = -\left(n-1-\frac{2p^2 + p + 2}{6p}\right)ln(U)
$$
$$
  U = \Lambda^{2/n}
$$
$$
  \Lambda = \frac{|S|^{n/2}}{(tr(S)/p)^{np/2}}
$$

And therefore:

$$
  U = \frac{p^p|S|}{(tr(S))^p}
$$

```{r}
S.det <- det(S)
S.tr <- psych::tr(S)
p <- dim(S)[2]
n <- dim(dados8)[1]

U = ((p^p)*S.det) / (S.tr^p)

Lambda <- (S.det^(n/2)) / ((S.tr/p)^(n*p/2))
Lambda^(2/n)

Ustar <- -(n - 1 - (2*p^2 + p + 2)/(6*p))*log(U)

## Chi sqrt degrees of freedom
chisqrt.degreesfreedom <- (p*(p+1))/2-1

```


What's the probability of U* to be greather than `r round(Ustar,3)`?

P(U* > `r round(Ustar,3)`)

```{r}
1 - pchisq(Ustar, chisqrt.degreesfreedom)

```

And therefore we reject H0. 
This means that there is at least one correlation that is significantly differente than 0.

### Now automated with R code

```{r}
mauchly.test(lm(dados8.scaled~1))
```

