---
title: "Computational Stats"
site: bookdown::bookdown_site
---

```{r include = FALSE}
datasetsDir=file.path(PROJHOME,"datasets")
```

To use the lab computers, the access credentials are:

> usr: em
> pwd: Emult*2018

# Lesson 1


```{r}
x <- 3+5

ls()
```

## Start by creating a vector


```{r}
y <- c(2,5,9,8)

y[1:3]

y[c(1,3)]

```

#### Get the elements 1,2,3 from the vector

```{r}
y[1:3]
```

#### Get the elements 1,3 from the vector

```{r}
y[c(1,3)]
```

#### Get an array from 0 to 1, with a 0.001 step

```{r}
y <- 1:1000/1000
y <- seq(0,1,0.001)
```

#### Which values are lower than 0.008?

```{r}
isValueLowerThan <- y < 0.008
y[isValueLowerThan]

idxs <- which(y<0.08)
y[idxs]
```

#### Creating objects by repetition

```{r}
colors <- c("amarelo","verde","vermelho","azul")

rep(colors, 5)
print("===")
rep(10,5)
```

## Now a Matrix!

```{r}
M <- matrix(1:9, ncol=3)
M
```

Transposing the Matrix
```{r}
t(M)
```

Accessing the Matrix
```{r}
M[1,2]
M[1,]
M[,2]
```

Matrix Operation
```{r}
M2 <- t(M)

M+M2 # valuewise add

M*M2 # valuewise multiplication

M%*%M2 # Matricial Multiplication

```

#### Joining Matrixes

Matrix Operation
```{r}

cbind(M,M2)

rbind(M,M2)

```

#### Inverting a matrix
```{r}

#solve(M) # M must not be singular 

```

## DataFrames

```{r}

y <- 1:10
y2 <- 11:20
y3 <- letters[1:10]

d1 <- data.frame(y,y2,y3)

d1
```


## Reading a Tab Separated File

```{r}
emp <- read.table(file.path(datasetsDir,"empresas.txt"), header=F)
knitr::kable(head(emp))
dim(emp)

names(emp) <- c("nome","n.socios","c.social","vmm","n.emp")
knitr::kable(head(emp))

emp$n.socios
emp[,2]
```

## Generating data

```{r}
set.seed(5)

emp$ant <- round(rnorm(dim(emp)[1],10,1))
```


## Getting insights

```{r}
summary(emp)

mean(emp$n.socios)

sd(emp$n.socios)

tapply(emp$vmm, emp$n.emp, mean) # vmm mean by number of employes
tapply(emp$vmm, emp$n.emp, sd) # vmm sd by number of employes
```

$$ \overline{X} = \frac{1}{N}\sum\limits_{i=1}^{N}X_{i} $$

\begin{equation} 
    \label{eq:std}
	S^2 = \frac{1}{N}\sum\limits_{i=1}^{N}(X_{i} - \overline{X})^2
\end{equation}


```{r}

table(emp$n.emp) #first line are values, second line is frequency

barplot(emp$n.emp) # each company is a bin in x label, y is the number of employees

barplot(table(emp$n.emp), xlab="#Employees", ylab="Frequecy", col="pink")

boxplot(emp$vmm,range=0,col="purple",horizontal=T,main="vmm")

boxplot(emp$vmm ~ emp$n.emp,range=0,col="purple",horizontal=T,main="vmm")

hist(emp$vmm)

hist(emp$vmm, freq=F)
lines(density(emp$vmm),col=2)


par(mfrow=c(1,2))
hist(emp$vmm)

hist(emp$vmm, freq=F)
lines(density(emp$vmm),col=2)
par(mfrow=c(1,1))

plot(emp$vmm,emp$n.socios,pch=16)

plot(emp)
```

## Lists

```{r}
uma.lista <- list(
  um.vector=1:10,
  uma.palavra="olá",
  uma.matrix=M,
  outra.lista=list(
    a="flor",
    b=rep(3,5)
  )
)

uma.lista["um.vector"]
uma.lista$um.vector
uma.lista[1]
```

## Functions

```{r}

desconto <- function(price, discount=25){
  #Discount is a number between 0 and 100
  #calcula o desconto de um preço
  newPrice <- price*(1-discount/100)
  discount <- price - newPrice
  list(
    novo.preco=newPrice,
    desconto=discount)
}

desconto(1000,20)
desconto(1000,25)

```

This is how you function
